---
title: "final_project_draft"
author: "Omer Shasman"
date: "5/3/2022"
output: pdf_document
---

```{r setup, include=FALSE}
# knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(echo = TRUE,fig.width = 12,fig.height = 4)
```


# Introduction
On Time Performance analysis of an airline network - This is an important metric for the airline that is calculated as the percentage of flights which are delayed by more than 14 minutes while the aircraft arrives at the gate. There are multiple reasons which contribute to the variation in OTP. An analysis of the OTP metric breaking it down into its individual components namely different delay and historical delays can provide insights into how the OTP for an airline can be managed by operational/process changes. The Department of Transport releases the flight level, On Time Performance data. This dataset also has various other factors which affect the Arrival Delay of a flight. An exploratory analysis of this data with the Arrival Delay as the response variable analyzed against different dimensions provided in the dataset can reveal several insights to improve the OTP of an Airline.

## What
As part of my Final Project, I am planning to use a subset of OTP data to perform analysis of delays on actual file arrivals focusing on one particular Station and Airline. Since airline operations are very complex, the arrival delays itself can be due to varying factors, like weather delay, carrier delays, security delays, Late aircraft delay...etc or any combinations of any of these in general. My focus is only on 2 types of delays so that I can minimize the complexities in data structures and limit any repeating processes or steps, and rather focus on how to manipulate and do analysis/inference with few variables. Hence I will be considering only 5 years data ranging from year 2014 till 2019 two types of delays "Weather Delays" and "Carrier Delays" 

## Why
I thought airline is an interesting business with lot of complex operation/data and business itself is most of us are familiar with. Also, with the time constraint we have, there were few sites like [Kaggle](https://www.kaggle.com/datasets/yuanyuwendymu/airline-delay-and-cancellation-data-2009-2018?select=2014.csv)  and [DOT On-Time_performance](https://www.transtats.bts.gov/databases.asp?f7owrp6_VQ=G&f7owrp6_Qr5p=cn55r0tr4%FDg4n8ry&Z1qr_VQF=D)

This data is presented as yearly file in csv format, I have to merge different years data using dply bind command to append rows at the end and build one file.

## How
Use RMarkdown and explore Rfunctions that can integrate some of the topics we learned in the class for flight arrival delay analysis. The following are steps which will be followed as part of the project

- Load data into R Markdown using R chunks
- Merge Data using ddplyr 
- Filter/melt/massage data using Tidy Data approach
- Use sampling strategy for identifying sample observations.
- Use Stats function to determine mean, median, IQR,...
- Regression - Find any co-relation between total flights arriving at a particualr airport and  delays to identify if it's the airport operational/capacity issue or not.
- Could hosted data and R Markdown interface. Pull data from AWS S3 buckets than loading from local machine.

# Body
This project perform analysis of flight arrival delays focusing on one particular Station and Airline.
Since airline delays are unavoidable there is always a chance that a particualr flight will be delayed. I think this analysis can be used to further study on why a particular delay happens and if the process/schedule/operations can be enhanced or refined to minimize the delay risk in future flights. 

# Packages Required
```{r message=TRUE, warning=FALSE, paged.print=TRUE}
library(knitr) ##for printing tables in R Markdown
library(dplyr) ##for data munging
library(ggplot2) ## for charts
library(infer) ## for rep_sample_n used for clustered sampling
```


```{r Load data into R Markdown using R chunks, message=FALSE, warning=FALSE, paged.print=FALSE}
library(readr)
flight.data.y2014 <- read_csv("Data/2014.csv")
# head(flight.data.y2014)

flight.data.y2015 <- read_csv("Data/2015.csv")
# head(flight.data.y2015)

flight.data.y2016 <- read_csv("Data/2016.csv")
# head(flight.data.y2016)

flight.data.y2017 <- read_csv("Data/2017.csv")
# head(flight.data.y2017)

flight.data.y2018 <- read_csv("Data/2018.csv")
# head(flight.data.y2018)

# Since this is a large dataset, sampling/manipualting on all the observations 
# is throwing memory error in my machine. So for ease of processing, I am 
# considering a subset of data with arrival station as MSP.

flight.data.y2014 <- flight.data.y2014[flight.data.y2014$DEST %in% 'MSP', ]
flight.data.y2015 <- flight.data.y2015[flight.data.y2015$DEST %in% 'MSP', ]
flight.data.y2016 <- flight.data.y2016[flight.data.y2016$DEST %in% 'MSP', ]
flight.data.y2017 <- flight.data.y2017[flight.data.y2017$DEST %in% 'MSP', ]
flight.data.y2018 <- flight.data.y2018[flight.data.y2018$DEST %in% 'MSP', ]
```

```{r Merge Data using ddplyr}
# Combine the 5 vectors(for each files) to a single vector
flight.data.y2014.y2018 <- 
  dplyr::bind_rows(flight.data.y2014,
                   flight.data.y2015,
                   flight.data.y2016,
                   flight.data.y2017,
                   flight.data.y2018,
                   )

# replace all na in valriables which we interested in to 0 for summary calculations.
flight.data.y2014.y2018$ARR_DELAY <- 
  flight.data.y2014.y2018$ARR_DELAY %>% 
  replace(is.na(.), 0)

flight.data.y2014.y2018$LATE_AIRCRAFT_DELAY <- 
  flight.data.y2014.y2018$LATE_AIRCRAFT_DELAY %>% 
  replace(is.na(.), 0)

flight.data.y2014.y2018$SECURITY_DELAY <- 
  flight.data.y2014.y2018$SECURITY_DELAY %>% 
  replace(is.na(.), 0)

flight.data.y2014.y2018$WEATHER_DELAY <- 
  flight.data.y2014.y2018$WEATHER_DELAY %>% 
  replace(is.na(.), 0)

flight.data.y2014.y2018$CARRIER_DELAY <- 
  flight.data.y2014.y2018$CARRIER_DELAY %>% 
  replace(is.na(.), 0)

hist(flight.data.y2014.y2018$FL_DATE, 
     flight.data.y2014.y2018$ARR_DELAY, 
     breaks = 100, 
     col='lightsteelblue3')

summary(flight.data.y2014.y2018$ARR_DELAY)

```

```{r demonstarte plot function of carrier delay greater than 60 mins against airtime of the flight}
flight.data.y2014.y2018.delay <- 
  flight.data.y2014.y2018[flight.data.y2014.y2018$CARRIER_DELAY > 60, ]

summary(flight.data.y2014.y2018.delay$CARRIER_DELAY)

plot(flight.data.y2014.y2018.delay$AIR_TIME, 
     flight.data.y2014.y2018.delay$CARRIER_DELAY, 
     col = "darkgreen")
```


# Topics From Class

## Topic 1:
R Markdown - I will be presenting the project in R Markdown and knit the file to a pdf document. Will be using R chunks to demonstrate and build the project components. 

## Topic 2:
GitHub - Will host the project in github repository for others to view my project components. 

## Topic 3:
Sampling strategies for an Observational study - Will be using sampling strategies - Simple random sampling, Strtified sampling, Cluster sampling and multistage sampling to group the data together by using different variables from the dataset and then use one of the sampling result to build topic#4 and 5.

```{r Sampling_Strategy - Simple sampling}
simple.sampling <- dplyr::sample_n(flight.data.y2014.y2018, 1000, replace=FALSE)
# View(simple.sampling)
simple.sampling
```

```{r Sampling_Strategy - Stratified sampling}
# Here I am making a cluster of where Airline code is the strata
DL <- flight.data.y2014.y2018[flight.data.y2014.y2018$OP_CARRIER %in% 'DL', ]
UA <- flight.data.y2014.y2018[flight.data.y2014.y2018$OP_CARRIER %in% 'UA',]
AA <- flight.data.y2014.y2018[flight.data.y2014.y2018$OP_CARRIER %in% 'AA',]
WN <- flight.data.y2014.y2018[flight.data.y2014.y2018$OP_CARRIER %in% 'WN',]

stratified.sampling <- dplyr::sample_n((UA), 1000, replace=FALSE)
dim(stratified.sampling)
```

```{r Sampling_Strategy - Clustered sampling}
#randomly choose 4 10 groups out of the n
clusters <- 
  sample(unique(flight.data.y2014.y2018$OP_CARRIER), size=10, replace=FALSE)

#define sample as all members who belong to one of the 10 operated carriers
clustered_by_op_carrier <- 
  flight.data.y2014.y2018[flight.data.y2014.y2018$OP_CARRIER %in% clusters, ]

#view how many observations came from each tour
table(clustered_by_op_carrier$OP_CARRIER)
clustered_by_op_carrier

```


## Topic 4:
Detailing Summary statistics ( Min. , 1st Qu., Median, Mean, 3rd Qu., Max.) of a variable and plotting graphs using ggplot2


```{r CARRIER_DELAY Years 2014 to 2028}
carrier_delay_by_op_carrier <- 
  clustered_by_op_carrier %>% 
  select(c(FL_DATE, OP_CARRIER, CARRIER_DELAY)) %>% 
  filter(OP_CARRIER == 'DL') %>%  filter(CARRIER_DELAY > 0) 

summary(carrier_delay_by_op_carrier$CARRIER_DELAY)
hist(carrier_delay_by_op_carrier$FL_DATE, 
     carrier_delay_by_op_carrier$CARRIER_DELAY, 
     breaks = 100, 
     col = "lightblue")

```

```{r CARRIER_DELAY - Months in 2016. }
carrier_delay_by_op_carrier_2016 <- 
  carrier_delay_by_op_carrier

carrier_delay_by_op_carrier_2016$YEAR <- 
  format(carrier_delay_by_op_carrier$FL_DATE, "%Y")

carrier_delay_by_op_carrier_2016 <- 
  carrier_delay_by_op_carrier_2016 %>% 
  filter(YEAR == '2016')

carrier_delay_by_op_carrier_2016$MONTH <- 
  format(carrier_delay_by_op_carrier_2016$FL_DATE, "%m")

hist(carrier_delay_by_op_carrier_2016$FL_DATE, 
     carrier_delay_by_op_carrier_2016$CARRIER_DELAY, 
     breaks =200, 
     col = "lightgreen")

```
```{r CARRIER_DELAY - Day in Aug2016}
# From the above plot it has been determined that August2016 has an increase in carrier delay. The following will plot by Days in Aug2016 to determine which day have an spike in Carrier Delay.
carrier_delay_by_op_carrier_2016_AUG <- 
  carrier_delay_by_op_carrier_2016

carrier_delay_by_op_carrier_2016_AUG <- 
  carrier_delay_by_op_carrier_2016_AUG %>% 
  filter(MONTH == '08')

carrier_delay_by_op_carrier_2016_AUG$DAY <- 
  format(carrier_delay_by_op_carrier_2016_AUG$FL_DATE, 
         "%d")

hist(carrier_delay_by_op_carrier_2016_AUG$FL_DATE, 
     carrier_delay_by_op_carrier_2016_AUG$CARRIER_DELAY, 
     breaks =40, 
     col = "orange")

summary(carrier_delay_by_op_carrier_2016_AUG$CARRIER_DELAY)
```
From the news during those days there was a system outage which caused this massive delay/cancellations.
[Delta System Outage - Aug2016](https://money.cnn.com/2016/08/08/news/companies/delta-system-outage-flights/)

## WEATHER DELAY
```{r WEATHER DELAY - Years 2014-2028}
# First select the sub-vectors which contains only he coulums we are interested in
weather_delay_by_op_carrier <- 
  clustered_by_op_carrier %>% 
  select(c(FL_DATE, 
           OP_CARRIER, 
           WEATHER_DELAY)) %>% 
  filter(OP_CARRIER == 'DL') %>%  
  filter(WEATHER_DELAY > 0
         ) 

hist(weather_delay_by_op_carrier$FL_DATE, 
     weather_delay_by_op_carrier$WEATHER_DELAY, 
     breaks = 100, 
     col = "lightblue")
```

```{r WEATHER DELAY - Months in 2016}
# In 2016 we have a increase in delay. Let's find out month by analysis to approximate on which month in 2016 we have this issue.
weather_delay_by_op_carrier_2016 <- 
  weather_delay_by_op_carrier

weather_delay_by_op_carrier_2016$YEAR <- 
  format(weather_delay_by_op_carrier$FL_DATE, "%Y")

weather_delay_by_op_carrier_2016 <- 
  weather_delay_by_op_carrier_2016 %>% 
  filter(YEAR == '2016')

weather_delay_by_op_carrier_2016$MONTH <- 
  format(weather_delay_by_op_carrier_2016$FL_DATE, "%m")

hist(weather_delay_by_op_carrier_2016$FL_DATE, 
     weather_delay_by_op_carrier_2016$weather_delay, 
     breaks =200, col = "lightgreen")

```
The above histogram shows that whether delays are massive in December month in MSP airport. This could be explained by winter storms related delays.


```{r WEATHER DELAY - Days in Dec2016}
weather_delay_by_op_carrier_2016

weather_delay_by_op_carrier_2016_DEC <- 
  weather_delay_by_op_carrier_2016 %>% 
  filter(MONTH == '12')

weather_delay_by_op_carrier_2016_DEC$DAY <- 
  format(weather_delay_by_op_carrier_2016_DEC$FL_DATE, "%d")

# weather_delay_by_op_carrier_2016_DEC

hist(as.numeric(weather_delay_by_op_carrier_2016_DEC$DAY), 
     weather_delay_by_op_carrier_2016_DEC$weather_delay, 
     breaks =30, col = "orange")

summary(weather_delay_by_op_carrier_2016_DEC$weather_delay)
```




```{r}
weather_delay_by_op_carrier_2014.gorupby <- 
  flight.data.y2014 %>% 
  select(c(FL_DATE, OP_CARRIER, WEATHER_DELAY)) %>% 
  filter(OP_CARRIER == 'DL') %>% 
  filter(WEATHER_DELAY > 0) %>% 
  group_by(as.numeric(format(FL_DATE, "%m"))) %>% 
  summarize(total_delayed=sum(WEATHER_DELAY)) %>% 
  mutate(year=2014)

weather_delay_by_op_carrier_2015.gorupby <- 
  flight.data.y2015 %>% 
  select(c(FL_DATE, OP_CARRIER, WEATHER_DELAY)) %>% 
  filter(OP_CARRIER == 'DL') %>% 
  filter(WEATHER_DELAY > 0) %>% 
  group_by(as.numeric(format(FL_DATE, "%m"))) %>% 
  summarize(total_delayed=sum(WEATHER_DELAY)) %>% 
  mutate(year=2015)

weather_delay_by_op_carrier_2016.gorupby <- 
  flight.data.y2016 %>% 
  select(c(FL_DATE, OP_CARRIER, WEATHER_DELAY)) %>% 
  filter(OP_CARRIER == 'DL') %>% 
  filter(WEATHER_DELAY > 0) %>% 
  group_by(as.numeric(format(FL_DATE, "%m"))) %>% 
  summarize(total_delayed=sum(WEATHER_DELAY)) %>% 
  mutate(year=2016)

weather_delay_by_op_carrier_2017.gorupby <- 
  flight.data.y2017 %>% 
  select(c(FL_DATE, OP_CARRIER, WEATHER_DELAY)) %>% 
  filter(OP_CARRIER == 'DL') %>% 
  filter(WEATHER_DELAY > 0) %>% 
  group_by(as.numeric(format(FL_DATE, "%m"))) %>% 
  summarize(total_delayed=sum(WEATHER_DELAY)) %>% 
  mutate(year=2017)

weather_delay_by_op_carrier_2018.gorupby <- 
  flight.data.y2018 %>% 
  select(c(FL_DATE, OP_CARRIER, WEATHER_DELAY)) %>% 
  filter(OP_CARRIER == 'DL') %>% 
  filter(WEATHER_DELAY > 0) %>% 
  group_by(as.numeric(format(FL_DATE, "%m"))) %>% 
  summarize(total_delayed=sum(WEATHER_DELAY)) %>% 
  mutate(year=2018)

month_Delay<-rbind(weather_delay_by_op_carrier_2014.gorupby, 
                   weather_delay_by_op_carrier_2015.gorupby, 
                   weather_delay_by_op_carrier_2016.gorupby, 
                   weather_delay_by_op_carrier_2017.gorupby, 
                   weather_delay_by_op_carrier_2018.gorupby)

ggplot(month_Delay, 
       aes(x = `as.numeric(format(FL_DATE, "%m"))`, 
           y = total_delayed, 
           color = factor(year), group = factor(year))) +
geom_line(linetype = 1) +
  labs(title="Total Flights Delayed across Months",y = 'Number of Delays',x = 'Months', fill='YEAR')
```
The above plot shows weather delays during 2014-2018 and it's clear that most of the weather related delays happens during year start/end. 


## Topic 5:
Regression (if an increase in number of schedules has any impact/variace on carrier delays).


```{r regression}
flight.data.y2018
carrier_delay_by_op_carrier_2018.totalFlights <- 
  flight.data.y2018 %>% 
  select(c(FL_DATE, OP_CARRIER, CARRIER_DELAY)) %>% 
  filter(OP_CARRIER == 'DL') %>% 
  group_by(FL_DATE) %>% count() %>% mutate(year=2018) 

carrier_delay_by_op_carrier_2018.carrierDelay <- 
  flight.data.y2018 %>% 
  select(c(FL_DATE, OP_CARRIER, CARRIER_DELAY)) %>% 
  filter(OP_CARRIER == 'DL') %>% 
  filter(CARRIER_DELAY >= 0) %>% 
  group_by(FL_DATE) %>% 
  summarize(total_delayed=sum(CARRIER_DELAY > 0)) %>% 
  mutate(year=2018)

dataset <- bind_cols(carrier_delay_by_op_carrier_2018.totalFlights,
                     carrier_delay_by_op_carrier_2018.carrierDelay)

linear_model <- lm(total_delayed ~ n, 
                   data=dataset)
summary(linear_model)

```
```{r regplot, dev='png'}
ggplot(dataset, aes(x=n, 
                    y=total_delayed)) +
  geom_point() +
  geom_smooth(method='lm', se=FALSE, col="red", size=1)
```


There appears to be have a linear relation between carrier delay and total flights. Since there is a linear relation between number of flights arrived and carrier delay, it could be due to airline related issue(like crew/pilot scheduling issue or some other operational issues.)


# Additional topic: Access to aws s3 bucket in RMarkdown. As a fun expiriment I am trying to store the generated file or plots by a chunk into aws s3. Still this work is in progress

```{r}
library("aws.s3")
Sys.setenv(
  "AWS_ACCESS_KEY_ID" = "AKIAUTK5NLVJF67UNMH5",
  "AWS_SECRET_ACCESS_KEY" = "",
  "AWS_DEFAULT_REGION" = "us-east-1"
)
```

```{r}
bucketlist()
```

```{r}
tempdir()
tempfile()
write.csv('writeup.pdf', file.path(tempdir(),"writeup.pdf"))
```

```{r}
# put_object(
#   file = file.path(tempdir(), writeup.pdf),
#   object = 'writeup.pdf',
#   bucket = 'seis631-finalproject'
# )
```



# Conclusion
I designed this project as a way to review some of the topics we learned in the class/homework/assignments to reinforce some topics learned and also as an opportunity to refer back some of the materials. Hence I thought of picking a variety of topics like sampling strategies, summary statistics, ANOVA and regressions will be the best approach and most I can get from this project. If I have more time, I would have included some more topics (like binom, dbinom, geom...etc distributions) and see if my dataset have variables that can fit these distributions. Given only a academic background in statistics almost almost 20 years ago, I think this subject has given me much learning experience in statistics and I appreciate how these topics are applicable to find solutions in reality.



